---
title: "Final Project Shiny App"
output: flexdashboard::flex_dashboard
runtime: shiny
---


```{r setup, include=FALSE}
library(flexdashboard)
library(shiny)
library(rmarkdown)
library(knitr)
library(Hmisc)
library(DT)
library(lubridate)
library(choroplethr)
library(choroplethrZip)
library(tidyverse)
library(data.table)
library(randomForest) #random forest algo
library(DataExplorer) #plot missing values
library(caret) #splitting dataset
library(Hmisc) #function cut2
library(dlookr) #find skewness
library(glmnet) #lasso
library(e1071) #skewness func
library(bit64)
library(corrplot)
library(rpart)
library(rpart.plot)
assignInNamespace("cedta.override", c(data.table:::cedta.override,"rmarkdown"), "data.table")
opts_chunk$set(echo = FALSE, comment="", warning = FALSE, message = FALSE, tidy.opts=list(width.cutoff=55), tidy = TRUE)
```

```{r load_data}
setwd("~/Columbia/AppliedDS/FinalProject/AppliedDataScienceFinalProject")#
all.data <- fread("Data/NYCRealEstateFullData.csv")
year <- substr(all.data$`SALE DATE`, 1, 4)
all.data$`SALE YEAR` <- year
sale_year <- unique(all.data$`SALE YEAR`)
```

```{r constants}
old.borough.name <- "BOROUGH"
borough.name <- "Fixed Borough"
neighborhood.name <- "NEIGHBORHOOD"
building.class.name <- "BUILDING CLASS CATEGORY"
tax.class.name <- "TAX CLASS AT PRESENT"
block.name <- "BLOCK"
lot.name <- "LOT"
easement.name <- "EASE-MENT"
building.class.present.name <- "BUILDING CLASS AT PRESENT"
address.name <- "ADDRESS"
apartment.number.name <- "APARTMENT NUMBER"
zip.name <- "ZIP CODE"
residential.name <- "RESIDENTIAL UNITS"
commercial.name <- "COMMERCIAL UNITS"
total.units.name <- "TOTAL UNITS"
land.square.feet.name <- "LAND SQUARE FEET"
gross.square.feet.name <- "GROSS SQUARE FEET"
year.built.name <- "YEAR BUILT"
tax.class.sale.name <- "TAX CLASS AT TIME OF SALE"
building.class.sale.name <- "BUILDING CLASS AT TIME OF SALE"
sale.price.name <- "SALE PRICE"
sale.date.name <- "SALE DATE"
sale.year.name <- "Sale Year"
sale.year<-"SALE YEAR"
sale.month <- "SALE MONTH"
log.price.name <- "Log Price"
building.class.first.letter <- "Building Class First Letter"
all.data <- all.data[, `Sale Year` := year(get(sale.date.name))]
all.data[, `Building Class First Letter` := substr(get(building.class.sale.name), 1, 1)]
dat <- all.data[get(sale.price.name) > 50000]
year.variables <- unique(all.data[, get(sale.year.name)])
first.letter.variables <- unique(all.data[, get(building.class.first.letter)])
residential.codes <- c("A", "B", "C", "D", "RR", "R1", "R2", "R3", "R4", "R6", "R7", "R8", "R9")
borough_name <- c("Manhattan", "Bronx", "Brooklyn", "Queens", "Staten Island")

residential.group.name <- "RESIDENTIAL UNITS GROUP"
commercial.group.name <- "COMMERCIAL UNITS GROUP"
total.units.group.name <- "TOTAL UNIT GROUP"
land.group.name <- "LAND SQUARE FEET GROUP"
gross.group.name <- "GROSS SQUARE FEET GROUP"

cuts.residential <- c(0, 1, 2, 3, 5)
cuts.commercial <- c(0, 1, 2, 3, 5)
cuts.total <- c(0, 1, 2, 3, 5)
cuts.land <- c(0, 100, 1000, 2000, 3000)
cuts.gross <- c(0, 100, 1000, 2000, 3000)

all.data[, eval(residential.group.name) := cut2(x = get(residential.name), cuts = cuts.residential)]
all.data[, eval(commercial.group.name) := cut2(x = get(commercial.name), cuts = cuts.commercial)]
all.data[, eval(total.units.group.name) := cut2(x = get(total.units.name), cuts = cuts.total)]
all.data[, eval(land.group.name) := cut2(x = get(land.square.feet.name), cuts = cuts.land)]
all.data[, eval(gross.group.name) := cut2(x = get(gross.square.feet.name), cuts = cuts.gross)]

unique.residential.group <- all.data[, unique(get(residential.group.name))]
unique.commercial.group <- all.data[, unique(get(commercial.group.name))]
unique.total.units.group <- all.data[, unique(get(total.units.group.name))]
unique.land.group <- all.data[, unique(get(land.group.name))]
unique.gross.group <- all.data[, unique(get(gross.group.name))]

all.data[, `Log Price` := log(get(sale.price.name))]

year.variables <- unique(all.data[, get(sale.year.name)])
first.letter.variables <- unique(all.data[, get(building.class.first.letter)])
residential.codes <- c("A", "B", "C", "D", "RR", "R1", "R2", "R3", "R4", "R6", "R7", "R8", "R9")
line.variables <- c(borough.name, building.class.first.letter)

model.inputs <- c(borough.name, land.square.feet.name, gross.square.feet.name, sale.year, total.units.name, year.built.name, neighborhood.name)
```

```{r data_cleaning}
all.data$`SALE DATE`<-ymd(as.Date(all.data$`SALE DATE`))    #conversion from factor to date format
all.data$`SALE YEAR`<-year(all.data$`SALE DATE`)   #extracting year from sale date
all.data$`SALE MONTH`<-month(all.data$`SALE DATE`) #extracting months from sale date
occurences<-all.data[, .N, eval(sale.month)] #evenly distributed

month.wise.sale<-all.data[,.("Mean Sale"=mean(`SALE PRICE`)),sale.month] 
monthly.sales<-merge(month.wise.sale,occurences,by = sale.month)
monthly.sales<-setorderv(monthly.sales,order=-1,cols="Mean Sale")

setDT(all.data)
all.data.tax.class.na<-all.data[is.na(`TAX CLASS AT PRESENT`),.SD] #20640 na values
#option 1: to remove these 20640 values from dataset as most of other variables like land sqr feet are also 0 and also 20640 missing values in building class at present also correspond to these values

#cleaning dataset
all.data.clean<-all.data[!is.na(all.data$`TAX CLASS AT PRESENT`),.SD]
all.data.clean<-all.data.clean[!is.na(`BUILDING CLASS CATEGORY`),.SD]
all.data.clean$`YEAR BUILT`[all.data.clean$`YEAR BUILT`== 0] <- NA

#making brackets for year built(grouping prop based on the year they were built in)
cuts.year <- c(1900,1925,1950,1975,2000)
all.data.clean[, eval(year.built.name) := cut2(x = get(year.built.name), cuts = cuts.year)]
#removing 0 values from sale price
all.data.clean$`SALE PRICE`[all.data.clean$`SALE PRICE`==0]<-NA
all.data.clean$`LAND SQUARE FEET`[all.data.clean$`LAND SQUARE FEET`==0]<-NA
all.data.clean$`GROSS SQUARE FEET`[all.data.clean$`GROSS SQUARE FEET`==0]<-NA

all.data.clean<-all.data.clean[!is.na(all.data.clean$`YEAR BUILT`),.SD]
all.data.clean<-all.data.clean[!is.na(all.data.clean$`SALE PRICE`),.SD]
all.data.clean<-all.data.clean[!is.na(all.data.clean$`LAND SQUARE FEET`),.SD]
all.data.clean<-all.data.clean[!is.na(all.data.clean$`GROSS SQUARE FEET`),.SD]
all.data.clean<-select(all.data.clean,c(-7,-10))

neigh<-all.data.clean[,.N,NEIGHBORHOOD]
setorderv(neigh,col="N",order=-1)

neigh.50<-neigh[1:50,]

neigh.price<-all.data.clean[,.('mean price'=mean(`SALE PRICE`)),NEIGHBORHOOD]
setorderv(neigh.price,cols="mean price",order=-1)

neigh.analysis<-merge(neigh,neigh.price,by = neighborhood.name)
setorderv(neigh.analysis,cols="mean price",order=-1)

all.data.clean$`SALE PRICE`=scale(all.data.clean$`SALE PRICE`)

a<-createDataPartition(all.data.clean$`SALE PRICE`,p = 0.70,groups =100,list=FALSE)
traina<-all.data.clean[a,]
testa<-all.data.clean[-a,]

neigh.analysis.50<-neigh.analysis[1:50]
selected.neighbors<-all.data.clean[get(neighborhood.name) %in% neigh.analysis.50$NEIGHBORHOOD]

#split
set.seed(100)
n<-createDataPartition(selected.neighbors$`SALE PRICE`,p = 0.70,groups =100,list=FALSE)
train.n<-selected.neighbors[n,]
test.n<-selected.neighbors[-n,]
```

```{r functions}
round.numerics <- function(x, digits){
  if(is.numeric(x)){
    x <- round(x = x, digits = digits)
  }
  return(x)
}

graph.choropleth <- function(data, year){
  keep_cols = c(zip.name, sale.price.name)
  data(zip.regions)
  zip.prices <- data[, ..keep_cols]
  zip.prices <- zip.prices[, mean(get(sale.price.name), na.rm = TRUE), by = zip.name]
  colnames(zip.prices) <- c("region", "value")
  zip.prices$value <- as.numeric(zip.prices$value)
  zip.prices$region <- as.character(zip.prices$region)
  zip.prices <- zip.prices[region %in% zip.regions$region,]
  zip.prices <- zip.prices[value > 0,]
  
  plot.title <- c(year, " Average Sale Price by Zip Code")
  plot.title <- paste(plot.title, collapse="")
  
  choro.graph <- zip_choropleth(zip.prices,
                 zip_zoom = zip.prices$region,
                 title       =  plot.title,
                 legend      = "Average Sale Price")
  return (choro.graph)
}

avg.sale.price.by <- function(data, by.column.names){
  mean.price <- data[, .(`Avg. Price` = mean(get(sale.price.name), na.rm=TRUE)), by = by.column.names]
  return (mean.price)
}

engagement.model <- function(dt, outcome.name, input.names){

  res <- fit.model(dt = dt, outcome.name = outcome.name, input.names = input.names)
  return(res)
}
fit.model <- function(dt, outcome.name, input.names, digits = 3){
  the.formula <- reduce.formula(dt = dt, outcome.name = outcome.name, input.names = input.names)

  mod <- lm(formula = the.formula, data = dt)
  mod.summary <- linear.regression.summary(lm.mod = mod, digits = digits)
  mod.summary.rounded <- mod.summary[, lapply(X = .SD, FUN = "round.numerics", digits = digits)]
  return(mod.summary.rounded)
}

linear.regression.summary <- function(lm.mod, digits = 3){
  library(data.table)
  lm.coefs <- as.data.table(summary(lm.mod)$coefficients, keep.rownames = TRUE)
  alpha = 0.05
  z <- qnorm(p = 1-alpha/2, mean = 0, sd = 1)
  lm.coefs[, Coef.Lower.95 := Estimate - z * `Std. Error`]
  lm.coefs[, Coef.Upper.95 := Estimate + z * `Std. Error`]
  return(lm.coefs)
}

create.formula <- function(outcome.name, input.names, input.patterns = NA, all.data.names = NA, return.as = "character"){
  
  variable.names.from.patterns <- c()
  if(!is.na(input.patterns[1]) & !is.na(all.data.names[1])){
    pattern <- paste(input.patterns, collapse = "|")
    variable.names.from.patterns <- all.data.names[grep(pattern = pattern, x = all.data.names)]
  }
  all.input.names <- unique(c(input.names, variable.names.from.patterns))
  all.input.names <- all.input.names[all.input.names != outcome.name]
  
  if(!is.na(all.data.names[1])){
    all.input.names <- all.input.names[all.input.names %in% all.data.names]
  }

  input.names.delineated <- sprintf("`%s`", all.input.names)
  the.formula <- sprintf("`%s` ~ %s", outcome.name, paste(input.names.delineated, collapse = "+"))
  
  if(return.as == "formula"){
    return(as.formula(the.formula))
  }
  if(return.as != "formula"){
    return(the.formula)
  }
}
reduce.formula <- function(dt, outcome.name, input.names, input.patterns = NA, max.input.categories = 20, max.outcome.categories.to.search = 4, return.as = "formula"){
  require(data.table)
  dt <- setDT(dt)
  
  if(!(outcome.name %in% names(dt))){
    return("Error:  outcome.name is not in names(dt).")
  }
  
  pattern.names <- list()
  if(!is.na(input.patterns[1])){
    for(i in 1:length(input.patterns)){
      pattern.names[[i]] <- names(dt)[grep(pattern = input.patterns[i], x = names(dt))]
    }
  }
  all.input.names <- c(input.names, as.character(pattern.names))
  
  num.outcome.categories <- dt[!is.na(get(outcome.name)), length(unique(get(outcome.name)))]
  
  if(num.outcome.categories <= max.outcome.categories.to.search){
    num.unique.tab <- dt[, lapply(X = .SD, FUN = function(x){return(length(unique(x[!is.na(x)])))}), .SDcols = input.names, by = outcome.name]
    min.categories.tab <- num.unique.tab[, lapply(X = .SD, FUN = "min"), .SDcols = input.names]
    
    reduced.inputs <- names(min.categories.tab)[min.categories.tab >= 2]
  }
  if(num.outcome.categories > max.outcome.categories.to.search){
    reduced.inputs <- all.input.names
  }
  
  the.formula <- create.formula(outcome.name = outcome.name, input.names = reduced.inputs, all.data.names = names(dt), input.patterns = NA, return.as = return.as)
  return(the.formula)
}

add.backtick <- function(x, include.backtick = "as.needed"){
  if(include.backtick == "all"){
    w <- 1:length(x)
  }
  if(include.backtick == "as.needed"){
    w <- grep(pattern = " ", x = x, fixed = TRUE)
  }  
  if(length(w) > 0){
    x[w] <- sprintf("`%s`", x[w])
  }

  return(x)
}
```

Introduction
=====================================  

For our final project, we decided to analyze New York City real estate data between 2003 and 2017. The data is linked here: https://www1.nyc.gov/site/finance/taxes/property-annualized-sales-update.page

Click on the tabs to see different reports.


Choropleth Map
=====================================

Column {data-width=500}
-------------------------------------

```{r choropleth_map_1}
inputPanel(
  selectInput(inputId="year_variable_1", label = "Select Year 1:", choices = sort(year.variables), selected = sort(year.variables)[length(year.variables)]),
  selectInput(inputId="building_class_type_1", label = "Select Building Class Type 1:", choices = first.letter.variables, selected = residential.codes, multiple = TRUE)
)
renderPlot({
  choropleth.dat <- all.data[get(sale.year.name) == input$year_variable_1 & get(building.class.first.letter) %in% input$building_class_type_1,]
  graph.choropleth(choropleth.dat, input$year_variable_1)
})
```

Column {data-width=500}
-------------------------------------

```{r choropleth_map_2}
inputPanel(
  selectInput(inputId="year_variable_2", label = "Select Year 2:", choices = sort(year.variables), selected = sort(year.variables)[length(year.variables)-1]),
  selectInput(inputId="building_class_type_2", label = "Select Building Class Type 2:", choices = first.letter.variables, selected = residential.codes, multiple = TRUE)
)
renderPlot({
  choropleth.dat <- all.data[get(sale.year.name) == input$year_variable_2 & get(building.class.first.letter) %in% input$building_class_type_2,]
  graph.choropleth(choropleth.dat, input$year_variable_2)
})
```

Time-Series 
=====================================

Row {data-width=500}
-------------------------------------

```{r line_graph}
inputPanel(
  selectInput(inputId="line_variable", label = "Select Variable to Color By:", choices = line.variables, selected = line.variables[1]),
  selectInput(inputId="line_building_class_type", label = "Select Building Class Type:", choices = first.letter.variables, selected = residential.codes, multiple = TRUE)
)
renderPlot({
  if (input$line_variable == building.class.first.letter){
    sub.dat <- all.data[get(building.class.first.letter) %in% input$line_building_class_type,]
    mean.price <- avg.sale.price.by(sub.dat, c(sale.year.name, input$line_variable))
    setorderv(x = mean.price, cols = "Sale Year", order = 1)
    mean.price <- mean.price %>% mutate(`Building Class First Letter` = forcats::fct_reorder2(`Building Class First Letter`, `Sale Year`, `Avg. Price`))
    
    price.plot <- ggplot(mean.price, aes(`Sale Year`, as.integer(`Avg. Price`), color = `Building Class First Letter`)) + 
                   geom_line(size = 1) + geom_point(aes(`Sale Year`,as.integer(`Avg. Price`))) +
                   xlab("Sale Year") + ylab("Avg. Price") + labs(color = "Building Class") +
                   scale_x_continuous(breaks = scales::pretty_breaks(length(unique(mean.price$`Sale Year`)))) +
                   ggtitle("NYC Avg. Residential Real Estate Price by Building Class & Year") + theme_minimal(base_size = 14)
    price.plot
  }
  else{
    sub.dat <- all.data
    mean.price <- avg.sale.price.by(sub.dat, c(sale.year.name, input$line_variable))
    
    mean.year.price <- avg.sale.price.by(dat, c(sale.year.name))
    mean.year.price$`Fixed Borough` <- "Overall"

    mean.price <- rbind(mean.price, mean.year.price)

    setorderv(x = mean.price, cols = "Sale Year", order = 1)
    mean.price <- mean.price %>% mutate(`Fixed Borough` = forcats::fct_reorder2(`Fixed Borough`, `Sale Year`, `Avg. Price`))
    
    price.plot <- ggplot(mean.price, aes(`Sale Year`, as.integer(`Avg. Price`), color = `Fixed Borough`)) + 
                   geom_line(size = 1) + geom_point(aes(`Sale Year`,as.integer(`Avg. Price`))) +
                   xlab("Sale Year") + ylab("Avg. Price") + labs(color = "Borough") +
                   scale_x_continuous(breaks = scales::pretty_breaks(length(unique(mean.price$`Sale Year`)))) +
                   scale_color_manual(values=c("#3399FF", "#000000", "#66CC00", "#FF6633", "#9966FF", "#0066CC")) +
                   ggtitle("NYC Avg. Residential Real Estate Price by Borough & Year")  + theme_minimal(base_size = 14)
    price.plot
  }
})
```

Building Classes & Boroughs
==================================

The description of all the building class codes can be found [here](https://www1.nyc.gov/assets/finance/jump/hlpbldgcode.html).

Row {data-width=500}
----------------------------------

```{r class_borough}
inputPanel(
  selectInput(inputId="bar_building_class_type", label = "Select Building Class Type:", choices = first.letter.variables, selected = residential.codes, multiple = TRUE))
renderPlot({
  summary.building.codes <- all.data[, .N, by = c(building.class.first.letter, borough.name)]
  summary.building.codes <- summary.building.codes[N > 250, ]
  summary.building.codes <- summary.building.codes[get(building.class.first.letter) %in% input$bar_building_class_type]
  
  code.plot <- ggplot(summary.building.codes, aes(x = reorder(`Building Class First Letter`, -1*`N`), y = `N`, fill = `Fixed Borough`)) +
             geom_bar(stat = 'identity', position="dodge") + xlab("Building Class Code") + ylab("Number of Sales") + labs(fill = "Borough") +
             scale_fill_manual(values=c("#FF6633", "#66CC00", "#3399FF", "#9966FF", "#0066CC")) +
             theme_minimal(base_size = 20)

  code.plot
})
```

Number of Sales by Borough
======================================


Row {data-height=500}
-------------------------------------

```{r respondents}
inputPanel(
  selectInput(inputId = "input_year", label = "Select Year:", 
              choices = c(sale_year, 'All'), selected = 'All')
)
renderPlot({
  
  a <- matrix(0, nrow = 5, ncol = 2)
  colnames(a) <- c('borough', 'number')
  a <- as.data.frame(a)
  a[, 1] <- borough_name
  
  if (input$input_year == 'All'){
    for (i in 1:5){
     sub <- all.data[all.data$BOROUGH == i, ]
     a[i, 2] <- nrow(sub)
    }
    number_borough <- ggplot(data = a, aes(x = reorder(borough, -1*number), y = number)) +
      geom_bar(stat="identity", width=0.5, color = "blue", fill = "#99CCFF") + 
      xlab('Borough') + theme_minimal(base_size = 20)
    return(number_borough)
  }else{
    for (i in 1:5){
       sub <- all.data[all.data$BOROUGH == i & all.data$`SALE YEAR` == input$input_year, ]
       a[i, 2] <- nrow(sub)
    }
    number_borough <- ggplot(data = a, aes(x = reorder(borough, -1*number), y = number)) +
      geom_bar(stat = "identity", width = 0.5, color = "blue", fill = "#99CCFF") + 
      xlab('Borough') + theme_minimal(base_size = 20)
    return(number_borough)
  }
  
})
```

Number of Sales by Year
====================================


Row {data-width=500}
-------------------------------------

```{r}
inputPanel(
  selectInput(inputId = "line_borough", label = "Select Borough:", 
              choices = c(borough_name, 'Overall'), selected = "Overall", multiple = TRUE)
)
renderPlot({
  
  all.sales <- all.data[, .N, by = sale.year.name]
  all.sales$`Fixed Borough` <- "Overall"
  
  borough.sales <- all.data[, .N, by = c(sale.year.name, borough.name)]
  
  all.sales <- rbind(borough.sales, all.sales)
  
  all.sales <- all.sales[get(borough.name) %in% input$line_borough,]
  setorderv(x = all.sales, cols = "Sale Year", order = 1)
  all.sales <- all.sales %>% mutate(`Fixed Borough` = forcats::fct_reorder2(`Fixed Borough`, `Sale Year`, N))
  
  number_year <- ggplot(data = all.sales, aes(x = `Sale Year`, y = N, color = `Fixed Borough`)) +
      geom_line(size = 1) + xlab("Sale Year") + ylab("Number of Sales") + labs(color = "Borough") +
      scale_x_continuous(breaks = scales::pretty_breaks(length(unique(all.sales$`Sale Year`)))) +
      theme_minimal(base_size = 14) +
      scale_color_manual(values = c("#000000", "#9966FF", "#66CC00", "#3399FF", "#0066CC", "#FF6633"))
  number_year
})
```

Linear Regression Model
==================================

Select the inputs you'd like to include in the linear regression model. Note that the dependent variable here is the scaled version of the price of each property.

Row {data-height = 300}
----------------------------------

```{r engagement_models}
inputPanel(
  selectInput(inputId='model_inputs', label = "Inputs to Model", choice = model.inputs, selected = model.inputs, multiple = TRUE)
)


renderDataTable({
  
  datatable(engagement.model(train.n, sale.price.name, input$model_inputs))
})

```
