---
title: "A Deep Dive Into New York Real Estate"
author: "Unique Team B"
date: "4/25/2019"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(data.table)
library(DT)
library(tidyverse)
library(extracat)
library(lubridate)
library(choroplethr)
library(choroplethrZip)
library(randomForest) #random forest algo
library(DataExplorer) #plot missing values
library(caret) #splitting dataset
library(Hmisc) #function cut2
library(dlookr) #find skewness
library(glmnet) #lasso
library(e1071) #skewness func
library(bit64)
library(corrplot)

setwd("~/Columbia/AppliedDS/FinalProject/AppliedDataScienceFinalProject")
all.data <- fread("Data/NYCRealEstateFullData.csv")
```

```{r functions, include = FALSE}
graph.choropleth <- function(data, year){
  keep_cols = c(zip.name, sale.price.name)
  data(zip.regions)
  zip.prices <- data[get(sale.year.name) == year, ..keep_cols]
  zip.prices <- zip.prices[, mean(get(sale.price.name), na.rm = TRUE), by = zip.name]
  colnames(zip.prices) <- c("region", "value")
  zip.prices$value <- as.numeric(zip.prices$value)
  zip.prices$region <- as.character(zip.prices$region)
  zip.prices <- zip.prices[region %in% zip.regions$region,]
  zip.prices <- zip.prices[value > 0,]
  
  plot.title <- c(year, " Average Sale Price by Zip Code")
  plot.title <- paste(plot.title, collapse="")
  
  choro.graph <- zip_choropleth(zip.prices,
                 zip_zoom = zip.prices$region,
                 title       =  plot.title,
                 legend      = "Average Sale Price")
  return (choro.graph)
}

avg.sale.price.by <- function(data, by.column.names){
  mean.price <- data[, .(`Avg. Price` = mean(get(sale.price.name), na.rm=TRUE)), by = by.column.names]
  return (mean.price)
}

## Data size function
set.size<-function(n,dat){
  the.rows<-sample(x=1:nrow(dat),size=n,replace = FALSE)
  return(dat[the.rows,])
}

linear.regression.summary <- function(lm.mod, digits = 3, alpha = 0.05) {
lm.coefs <- as.data.table(summary(lm.mod)$coefficients,
keep.rownames = TRUE)
setnames(x = lm.coefs, old = "rn", new = "Variable")
z <- qnorm(p = 1 - alpha/2, mean = 0, sd = 1)
lm.coefs[, Coef.Lower.95 := Estimate - z * `Std. Error`]
lm.coefs[, Coef.Upper.95 := Estimate + z * `Std. Error`]
return(lm.coefs[])
}

```


```{r constants}
old.borough.name <- "BOROUGH"
borough.name <- "Fixed Borough"
neighborhood.name <- "NEIGHBORHOOD"
building.class.name <- "BUILDING CLASS CATEGORY"
tax.class.name <- "TAX CLASS AT PRESENT"
block.name <- "BLOCK"
lot.name <- "LOT"
easement.name <- "EASE-MENT"
building.class.present.name <- "BUILDING CLASS AT PRESENT"
address.name <- "ADDRESS"
apartment.number.name <- "APARTMENT NUMBER"
zip.name <- "ZIP CODE"
residential.name <- "RESIDENTIAL UNITS"
commercial.name <- "COMMERCIAL UNITS"
total.units.name <- "TOTAL UNITS"
land.square.feet.name <- "LAND SQUARE FEET"
gross.square.feet.name <- "GROSS SQUARE FEET"
year.built.name <- "YEAR BUILT"
tax.class.sale.name <- "TAX CLASS AT TIME OF SALE"
building.class.sale.name <- "BUILDING CLASS AT TIME OF SALE"
sale.price.name <- "SALE PRICE"
sale.date.name <- "SALE DATE"
sale.year.name <- "Sale Year"
log.price.name <- "Log Price"
sale.month<-"SALE MONTH"
sale.year<-"SALE YEAR"
building.class.first.letter <- "Building Class First Letter"
year.built.name <- "YEAR BUILT"
borough_name <- c("Manhattan", "Bronx", "Brooklyn", "Queens", "Staten Island")

residential.group.name <- "RESIDENTIAL UNITS GROUP"
commercial.group.name <- "COMMERCIAL UNITS GROUP"
total.units.group.name <- "TOTAL UNIT GROUP"
land.group.name <- "LAND SQUARE FEET GROUP"
gross.group.name <- "GROSS SQUARE FEET GROUP"
cuts.total <- c(0, 1, 2, 3, 5)
cuts.land <- c(0, 100, 1000, 2000, 3000)
all.data[, eval(total.units.group.name) := cut2(x = get(total.units.name), cuts = cuts.total)]
all.data[, eval(land.group.name) := cut2(x = get(land.square.feet.name), cuts = cuts.land)]
unique.total.units.group <- all.data[, unique(get(total.units.group.name))]
unique.land.group <- all.data[, unique(get(land.group.name))]
scaled.price.name <- "Scaled Price"
scaled.total.units <- "Scaled Total Units"
all.data[, eval(scaled.price.name) := scale(x = get(sale.price.name))]
all.data[, eval(scaled.total.units) := scale(x = get(total.units.name))]
```

## Introduction to The Problem

- New York real estate data from 2003 to 2017
- Dataset can be found here: https://www1.nyc.gov/site/finance/taxes/property-annualized-sales-update.page
- Key questions: 
  + What variables are most influential in determining sale price?
  + Is location really as big a determinant as we might think?
  + What kind of real estate is being sold? 


## Description of the Data

- The dataset has `r all.data[, .N]` rows and `r length(colnames(all.data))` columns
- The major important outcome is the sale price
- Some informative predictors are location (borough and neighborhood) and year of sale

## Missing Data
```{r missing data exploration, include = FALSE}
plot_missing(all.data) #100% missing values for Ease.ment, 78% missing values in apartment number (can be removed from analysis)
```

## Unexpected Values in Dataset (Challenge #1)

-  Approach to tackle Challenge 1
- 30% of properties are not priced 

- reason could be change/transfer of ownership (quit claim deed))

- removing Year Built Values and Sale prices which are 0 for analysis purposes
```{r cleansing data}
setDT(all.data)
all.data.tax.class.na<-all.data[is.na(`TAX CLASS AT PRESENT`),.SD] #20640 na values
#option 1: to remove these 20640 values from dataset as most of other variables like land sqr feet are also 0 and also 20640 missing values in building class at present also correspond to these values

#cleaning dataset
all.data.clean<-all.data[!is.na(all.data$`TAX CLASS AT PRESENT`),.SD]
all.data.clean<-all.data.clean[!is.na(`BUILDING CLASS CATEGORY`),.SD]
all.data.clean$`YEAR BUILT`[all.data.clean$`YEAR BUILT`== 0] <- NA

#making brackets for year built(grouping prop based on the year they were built in)
cuts.year <- c(1900,1925,1950,1975,2000)
all.data.clean[, eval(year.built.name) := cut2(x = get(year.built.name), cuts = cuts.year)]
#removing 0 values from sale price
all.data.clean$`SALE PRICE`[all.data.clean$`SALE PRICE`==0]<-NA
all.data.clean$`LAND SQUARE FEET`[all.data.clean$`LAND SQUARE FEET`==0]<-NA
all.data.clean$`GROSS SQUARE FEET`[all.data.clean$`GROSS SQUARE FEET`==0]<-NA

plot_missing(all.data.clean)
find_skewness(all.data.clean) #18 i.e sale price is skewed
#transform sale price to reduce skewness in data
#all.data.clean$SALE.PRICE=transform(all.data.clean$SALE.PRICE)
#all.data.clean$LAND.SQUARE.FEET=transform(all.data.clean$LAND.SQUARE.FEET)
#all.data.clean$GROSS.SQUARE.FEET=transform(all.data.clean$GROSS.SQUARE.FEET)

all.data.clean<-all.data.clean[!is.na(all.data.clean$`YEAR BUILT`),.SD]
all.data.clean<-all.data.clean[!is.na(all.data.clean$`SALE PRICE`),.SD]
all.data.clean<-all.data.clean[!is.na(all.data.clean$`LAND SQUARE FEET`),.SD]
all.data.clean<-all.data.clean[!is.na(all.data.clean$`GROSS SQUARE FEET`),.SD]
all.data.clean<-select(all.data.clean,c(-7,-10))
plot_missing(all.data.clean)
```

## Occurences by Borough

```{r fix_borough,echo=FALSE}
setDT(all.data)
borough.wise.occurences<-all.data[,.N,by = `Fixed Borough`]
borough.wise.occurences<-setorderv(borough.wise.occurences,order=-1,cols="N")
datatable(borough.wise.occurences)
```

## Average Prices Through the Years

```{r avg_sale_price_by_year, echo = FALSE}
all.data[, `Log Price` := log(get(sale.price.name))]
all.data[, `Sale Year` := year(get(sale.date.name))]
all.data[, `Building Class First Letter` := substr(get(building.class.sale.name), 1, 1)]
dat <- all.data[get(sale.price.name) > 50000]
residential.codes <- c("A", "B", "C", "D", "RR", "R1", "R2", "R3", "R4", "R6", "R7", "R8", "R9")
residential.properties <- dat[get(building.class.first.letter) %in% residential.codes,]

mean.year.price <- avg.sale.price.by(dat, c(sale.year.name))
setorderv(x = mean.year.price, cols = "Sale Year", order = 1)

year.price.plot <- ggplot(mean.year.price, aes(`Sale Year`, as.integer(`Avg. Price`))) + 
                   geom_line(size = 1) + geom_point(aes(`Sale Year`,as.integer(`Avg. Price`))) +
                   xlab("Sale Year") + ylab("Avg. Price") +
                   scale_y_continuous(name="Av. Price", limits=c(500000, 2000000)) +
                   scale_x_continuous(breaks = scales::pretty_breaks(length(mean.year.price$`Sale Year`))) +
                   ggtitle("NYC Avg. Real Estate Price by Year") 

year.price.plot
```

## Average Prices Through the Boroughs

```{r avg_sale_price_borough, echo = FALSE, fig.width = 10}
mean.year.borough.price <- avg.sale.price.by(dat, c(sale.year.name, borough.name))
setorderv(x = mean.year.borough.price, cols = "Sale Year", order = 1)
mean.year.borough.price <- mean.year.borough.price %>% mutate(`Fixed Borough` = forcats::fct_reorder2(`Fixed Borough`, `Sale Year`, `Avg. Price`))

year.borough.price.plot <- ggplot(mean.year.borough.price, aes(`Sale Year`, as.integer(`Avg. Price`), color = `Fixed Borough`)) + 
                   geom_line(size = 1) + geom_point(aes(`Sale Year`,as.integer(`Avg. Price`))) +
                   xlab("Sale Year") + ylab("Avg. Price") + labs(color = "Borough") +
                   scale_x_continuous(breaks = scales::pretty_breaks(length(unique(mean.year.borough.price$`Sale Year`)))) +
                   ggtitle("NYC Avg. Real Estate Price by Borough and Year") 

year.borough.price.plot
```


## Plot of Number of House Sales in Each Year
```{r echo = F}
all.data[, `SALE YEAR` := year(get(sale.date.name))]
number_sale_year <- as.data.frame(all.data[, (.N), by = `SALE YEAR`])
for (i in 1:5){
  a <- as.data.frame(all.data[BOROUGH == i, (.N), by = `SALE YEAR`])
  number_sale_year <- rbind(number_sale_year, a)
}
number_sale_year$Region <- c(rep('Overall', 15), rep('Manhattan', 15), rep('Bronx', 15),
                              rep('Brooklyn', 15), rep('Queens', 15), rep('Staten Island', 15))
colnames(number_sale_year)[2] <- 'Number of Sale'
number_year <- ggplot(data = number_sale_year, aes(x = `SALE YEAR`, y = `Number of Sale`, color = Region, group = Region)) +
  geom_line(size = 1) + 
  xlab('Year') +
  ylab('Number of Sale') +
  scale_fill_brewer(palette = 'Accent')
number_year
```


## Analyzing Sale Prices with respect to Month of Sale

```{r fix sale.date}
all.data$`SALE DATE`<-ymd(all.data$`SALE DATE`)    #conversion from factor to date format
all.data$`SALE YEAR`<-year(all.data$`SALE DATE`)   #extracting year from sale date
all.data$`SALE MONTH`<-month(all.data$`SALE DATE`) #extracting months from sale date 
occurences<-all.data[, .N, eval(sale.month)] #evenly distributed

month.wise.sale<-all.data[,.("Mean Sale"=mean(`SALE PRICE`)),sale.month] 
monthly.sales<-merge(month.wise.sale,occurences,by = sale.month)
monthly.sales<-setorderv(monthly.sales,order=-1,cols="Mean Sale")
datatable(monthly.sales)
```

## The scaled price over total units and land square feet with the original data

```{r scatter_unit_square.feet_1, echo = FALSE, cache = TRUE}
ggplot(data = all.data) +
  geom_point(mapping = aes(x = get(total.units.name), y = get(scaled.price.name))) +
  labs(title = "Scaled NYC Real Estate Price by Total Units", x = "Total Units", y = "Scaled Sale Price")
ggplot(data = all.data) +
  geom_point(mapping = aes(x = get(land.square.feet.name), y = get(scaled.price.name))) +
  labs(title = "Scaled NYC Real Estate Price by Land Square feet", x = "Land Square Feet", y = "Scaled Sale Price")
```

## The scaled price over total units and land square feet with the subset of data

```{r scatter_unit_square.feet_2, echo = FALSE, cache = TRUE}
sub.data.unit.feet <- all.data[get(total.units.name) > 0 & get(total.units.name) < 5000 & get(land.square.feet.name) > 0 & get(land.square.feet.name) < 50000000 & get(sale.price.name) > 50000 & get(sale.price.name) < 5000000, ]
ggplot(data = sub.data.unit.feet) +
  geom_point(mapping = aes(x = get(total.units.name), y = get(scaled.price.name))) +
  labs(title = "Scaled NYC Real Estate Price by Total Units", x = "Total Units", y = "Scaled Sale Price")
ggplot(data = sub.data.unit.feet) +
  geom_point(mapping = aes(x = get(land.square.feet.name), y = get(scaled.price.name))) +
  labs(title = "Scaled NYC Real Estate Price by Land Square feet", x = "Land Square Feet", y = "Scaled Sale Price")
```

## Cut total.units and land.square.feet into groups

```{r barplot_unit_square.feet_num, echo = FALSE, warning = FALSE}
unit.Num <- all.data[, .N, by = total.units.group.name]
ggplot(data = unit.Num, mapping = aes(x = get(total.units.group.name), y = N)) +
  geom_bar(stat = "identity", color = "blue", fill = "#99CCFF") +
  labs(x = "Group of Total Units", y = "Number of Cases in each Group")
land.Num <- all.data[, .N, by = land.group.name]
ggplot(data = land.Num, mapping = aes(x = get(land.group.name), y = N)) +
  geom_bar(stat = "identity", color = "blue", fill = "#99CCFF") + 
  labs(x = "Group of Land Square Feet", y = "Number of Cases in each Group") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Boxplot of the logarithm of price in different groups

```{r boxplot_unit_square.feet_price, warning = FALSE}
ggplot(data = all.data, aes(x = get(total.units.group.name), y = get(log.price.name))) +
  geom_boxplot() +
  labs(title = "Logarithm of NYC Real Estate Price by Total Units", x = "Group of Total Units", y = "Log Sale Price")
ggplot(data = all.data, aes(x = get(land.group.name), y = get(log.price.name))) +
  geom_boxplot() +
  labs(title = "Logarithm of NYC Real Estate Price by Land Square Feet", x = "Group of Land Square Feet", y = "Log Sale Price") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## Location, Location, Location

```{r choropleth_all, echo = FALSE}
all.data[, `Log Price` := log(get(sale.price.name))]
all.data[, `Sale Year` := year(get(sale.date.name))]
all.data[, `Building Class First Letter` := substr(get(building.class.sale.name), 1, 1)]
dat <- all.data[get(sale.price.name) > 50000]
residential.codes <- c("A", "B", "C", "D", "RR", "R1", "R2", "R3", "R4", "R6", "R7", "R8", "R9")
residential.properties <- dat[get(building.class.first.letter) %in% residential.codes,]
graph.choropleth(dat, 2017)
```

## Residential Properties

```{r choropleth_res}
graph.choropleth(residential.properties, 2017)
```

## Plot of Price per Foot in New York City From Region Aspect
```{r echo = F}
year <- substr(all.data$`SALE DATE`, 1, 4)
all.data$`SALE YEAR` <- year
sub <- all.data[all.data$`SALE PRICE` != 0 & all.data$`GROSS SQUARE FEET` != 0 & all.data$`SALE PRICE` > 500000, ]
sub$`PRICE PER FEET` <- sub$`SALE PRICE`/sub$`GROSS SQUARE FEET`
median_price_feet_2 <- as.data.frame(sub[, .(price = median(x = `PRICE PER FEET`, na.rm = T)), keyby = BOROUGH])
mean_price_feet_2 <- as.data.frame(sub[, .(price = mean(x = `PRICE PER FEET`, na.rm = T)), keyby = BOROUGH])
price_feet_2 <- rbind(median_price_feet_2, mean_price_feet_2)
price_feet_2$BOROUGH <- rep(borough_name, 2)
price_feet_2$Type <- c(rep('median', 5), rep('mean', 5))
price_feet_2$price <- round(price_feet_2$price, 0)
price_borough <- ggplot(data = price_feet_2, mapping = aes(x = factor(BOROUGH),y = price, fill = Type)) +
  geom_bar(stat = "identity", position = 'dodge') + 
  scale_fill_manual(values = c('steelblue', 'orange')) +
  xlab('Borough') +
  ylab('Price Per Square Foot') +
  geom_text(aes(label = price), vjust = 1.5, colour = "white", position = position_dodge(.9), size = 4)
price_borough
```

## Plot of Price per Foot in New York City From Time Aspect
```{r echo = F}
price_feet_2_time <- as.data.frame(sub[, .(price = median(x = `PRICE PER FEET`, na.rm = T)), keyby = `SALE YEAR`])
for (i in 1:5){
  a <- sub[BOROUGH == i, .(price = median(x = `PRICE PER FEET`, na.rm = T)), keyby = `SALE YEAR`]
  price_feet_2_time <- rbind(price_feet_2_time, a)
}
price_feet_2_time$Region <- c(rep('Overall', 15), rep('Manhattan', 15), rep('Bronx', 15),
                              rep('Brooklyn', 15), rep('Queens', 15), rep('Staten Island', 15))
price_year <- ggplot(data = price_feet_2_time, aes(x = `SALE YEAR`, y = price, color = Region, group = Region)) +
  geom_line(size = 1) + 
  xlab('Year') +
  ylab('Price per Square Foot') +
  scale_fill_brewer(palette = 'Accent')
price_year
```

## Correlations

- Gross Square Feet and Sale price are correlated
- However, Land square feet and Sale price show weak correlation
- Total Units and Sale price also show a moderate correlation

```{r correlations}
m<-cor(all.data.clean$`GROSS SQUARE FEET`,all.data.clean$`SALE PRICE`) #0.47
n<-cor(all.data.clean$`LAND SQUARE FEET`,all.data.clean$`SALE PRICE`)  #0.09
o<-cor(all.data.clean$`TOTAL UNITS`,all.data.clean$`SALE PRICE`)

correlations<-all.data.clean[,.("Land Square Feet and Sale Price"=n,"Gross Square Feet and Sale Price"=m,"Total Units and Sale Price"=o)]
correlations<-round(correlations,3)
datatable(correlations)

```

```{r scaling the prices}
all.data.clean$`SALE PRICE`=scale(all.data.clean$`SALE PRICE`)
```

```{r one family dwellings only}
setDT(all.data.clean)
all.data.clean$`SALE YEAR` <- year(all.data.clean$`SALE DATE`)
one.family.dwellings<-all.data.clean[`BUILDING CLASS CATEGORY`=="01  ONE FAMILY DWELLINGS",.SD]
one.family.dwellings.2<-all.data.clean[`BUILDING CLASS CATEGORY`=="01 ONE FAMILY DWELLINGS",.SD]
one.family<-rbind(one.family.dwellings,one.family.dwellings.2)
#one.family$`BUILDING CLASS CATEGORY`<-droplevels(one.family$`BUILDING CLASS CATEGORY`)
```


```{r split into train and test}
library(caret)
a<-createDataPartition(one.family$`SALE PRICE`,p = 0.70,groups =100,list=FALSE)
train<-one.family[a,]
test<-one.family[-a,]
```

```{r using 5000 sample size for model training}
train.5000<-set.size(5000,train)
```

## Linear Regression Model with Selected Variables

```{r linear regression}
mod.lm<-lm(`SALE PRICE`~`Fixed Borough`+`LAND SQUARE FEET`+`GROSS SQUARE FEET`+`SALE YEAR`+`TOTAL UNITS`+`YEAR BUILT`,data=train)
summary(mod.lm)
pred.lm<-predict(mod.lm,newdata=test)
rmse.lm = sqrt(mean((pred.lm-test$`SALE PRICE`)^2)) #rmse.lm 0.0488, R sqr: 0.49 

#linear.regression.summary(mod.lm)

model.linear<-as.data.frame(linear.regression.summary(mod.lm))
model.linear<-model.linear[,c(1,2,5)]
model.linear.round<-round(model.linear[,-1],3)
model.linear.final<-cbind(model.linear$Variable,model.linear.round)
datatable(model.linear.final)
```

## Analyzing Neighborhoods in terms of their Sale Prices 

```{r}
neigh<-all.data.clean[,.N,NEIGHBORHOOD]
setorderv(neigh,col="N",order=-1)

neigh.50<-neigh[1:50,]

neigh.price<-all.data.clean[,.('mean price'=mean(`SALE PRICE`)),NEIGHBORHOOD]
setorderv(neigh.price,cols="mean price",order=-1)

neigh.analysis<-merge(neigh,neigh.price,by = neighborhood.name)
setorderv(neigh.analysis,cols="mean price",order=-1)
datatable(neigh.analysis)
```



## Implementation

- Many possible building types in the data, from one-family homes to airport terminals
- Building classes generally organized by first letter in "Building Class at Time of Sale" column

```{r implementation, echo = TRUE}
all.data[, `Building Class First Letter` := substr(get(building.class.sale.name), 1, 1)]
```

## Limitations and Uncertainties

- Many rows of the dataset have prices that don't make sense (ie. 0, 10, 100, etc.)
- It is never clear if a buyer bought the whole building, a few units, or only a single unit in the building 
- Several building class types so a user should be aware of the difference in price range

